---
title: "Squad age profile"
author: "Vamsi Nath"
date: "7/25/2019"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300, fig.width=7)
```

## Arsenal Squad Age-Profile analysis

This is small analytics exercise to examine the distribution of minutes played based on the age of the players.

I got inspiration for this from the excellent [article](http://petermckeever.com/2019/04/creating-squad-age-profiles/) by @petermckeever.

His work was done in Python but I decided to try and re-create it using R.

We will begin by scraping the player-ages for the squad from ESPNFC. One thing I noticed was that they separated the goalie stats from the rest of the squad so we will have to scrape two separate tables and combine the data.

First, we'll load our library dependencies and specify the squad page URL we will be using.

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(lubridate)
library(stringr)
library(scales)

url <- "https://www.espn.com/soccer/team/squad/_/id/359/league/ENG.1/season/2018"

```

Next, we will scrape the goalie information. I had to check the page source-code to determine the correct table-indexes.

```{r}
goalie <- read_html(url) %>% 
  html_nodes("table") %>% 
  magrittr::extract2(3) %>% 
  html_table(fill = TRUE,  header = TRUE, trim = TRUE)
  
```

Now, we'll extract the rest of the squad details from the 2nd table.

```{r}
outfield <- read_html(url) %>% 
  html_nodes("table") %>% 
  magrittr::extract2(6) %>% 
  html_table(fill = TRUE,  header = TRUE, trim = TRUE)

```

Let's combine both sets into a single dataframe and take a peek.

```{r}
squad <- bind_rows(goalie, outfield)

```

Let's continue by keeping only our columns of interest.

```{r}
squad <- squad %>% 
  select(
    Name, 
    Position = POS, 
    Age, 
    HT, 
    WT, 
    NAT
    )

```

We can also see we need to do some cleanup on the player names which annoyingly have their number appended at the end. We can take care of this easily though with a string replace using a regular expression. Let's do a quick sanity-check before we make the update to our dataframe.

```{r}
squad %>% mutate(Name = str_replace(Name, "\\d+$", "")) %>% pull(Name)

```

OK, we can see the player-numbers were correctly removed from the Names so we'll go ahead and perform this cleaning step on our dataframe. While we're at it, let's also remove the "lbs" label from the WT column and convert it to an integer.

```{r}
squad <- squad %>% 
  mutate(Name = str_replace(Name, "\\d+$", "")) %>% 
  mutate(WT = as.integer(str_replace(WT, "lbs", ""))) 
  
```

Next, let's convert our HT values to a standard of inches. We'll have to separate out our height-feet and height-inches to calculate total height in inches.

```{r}
(squad <- squad %>% separate(HT, c("feet","inches","drop")) %>% 
  select(-drop) %>% 
  mutate_at(c("feet","inches"), as.integer) %>% 
  mutate(HT = (feet*12) + inches) %>% 
  select(Name, Position, Age, HT, WT, NAT) 
)

```

For our final cleaning step, we will convert the POS column which holds the player's position to a factor and change the level names.

```{r}
squad <- squad %>% 
  mutate(Position = as.factor(Position))

levels(squad$Position) <- c("Defense","Forward","Goalie","Midfield")
  
```

Now, we'll load our player-ratings data that was scraped from [WhoScored.com](http://whoscored.com)

```{r}
ratings <- readRDS("arsenal_whoscored_player_ratings.rds") %>% as_tibble
```

We'll aggregate to the player-level taking the average of their rating and calculate total minutes played.

```{r}
player_agg <- ratings %>% 
  group_by(player_name) %>% 
  summarise(
    total_mins = sum(minutes_played),
    avg_rating = mean(rating)
            ) %>% 
  arrange(desc(total_mins))
```

Now, we'll combine our aggregated player dataframe with our squad profile. This will serve as our datasource for our viz. I did have to make edits to Aubameyang's and Sokratis's names so that their data would align with our ratings data.

```{r}
squad[27,"Name"] = "Pierre Emerick-Aubameyang"
squad[4, "Name"] = "Sokratis"

afc <- inner_join(squad, player_agg, by=c("Name"="player_name"))
```

Now, let's produce our viz.

```{r}
theme_set(theme_light())

afc %>% 
  ggplot(aes(x = Age, y = total_mins/(38*90), label = Name)) +
  geom_point(aes(color = Position), size = 3.0, alpha = 0.9) +
  geom_text(size = 3, vjust = "middle", hjust = "left", nudge_x = 0.3) +
  scale_color_manual(values = c("black","blue","orange","maroon")) +
  scale_x_continuous(breaks = seq(18,40, by=2)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "% of Total EPL Minutes")

```

We can see the graph looks a bit messy with some overlapping names. We can use the `check_overlap` property to help deal with this, but I found it removed several key players from the graph. 

Another option to clean things up is by showing only the last-name. We'll use the handy tidyr `extract` function to accomplish this.

```{r}
afc <- extract(afc, Name, c("fname","lname"), "([^ ]+)(.*)", remove = FALSE)

afc[4, "lname"] = "Sokratis"



p <- afc %>% ggplot(aes(x = Age, y = total_mins/(38*90), label = lname))

p + 
  geom_point(aes(color = Position), size = 3.0, alpha = 0.9) +
  geom_text(size = 3, vjust = "top", hjust = "middle", nudge_y = -0.02, nudge_x = 0.5) +
  scale_color_manual(values = c("black","blue","orange","maroon")) +
  scale_x_continuous(breaks = seq(18,40, by=2)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "% Played of Total EPL Minutes")

```


This looks good but let's go one-step farther by adding a shaded region showing the peak age-bracket which we'll define as from 25 - 30 years old.

```{r}
p <- p +
  geom_rect(fill = "green", alpha = 0.01, aes(xmin=25, xmax=30, ymin=0, ymax=1)) +
  geom_point(aes(color = Position), size = 3.0, alpha = 0.9) +
  geom_text(size = 3, vjust = "top", hjust = "middle", nudge_y = -0.02, nudge_x = 0.5) +
  scale_color_manual(values = c("black","blue","orange","maroon")) +
  scale_x_continuous(breaks = seq(18,40, by=2)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "% Played of Total EPL Minutes")

p
```

we'll wrap up by adding a title and some annotations.

```{r}
p + labs(
  title = "Arsenal | Squad Age Profile Analysis",
  subtitle = "Premier League, 2018/19",
  caption = "Data from Whoscored.com/ESPN"
) +
  annotate("text", x = 28, y = 0.95, label = "Peak Years", size = 4.0) +
  annotate("text", x = 36.5, y = 0.99, label = "@DesiGoonerMD", size = 2.5)
```